---
## ---- Submission metadata (The VLDB Journal) ------------------------
## The active build target is `html_document` (development build); the
## submission build (`make vldb`, Springer svjour3 twocolumn) renders a
## copy of this file with paper/vldb/vldbj-template.tex. Title and
## abstract below are the single source for both builds; authors,
## institutes, and keywords for the submission build live in the
## template (keep in sync).
## --------------------------------------------------------------------

title: "FbSQL: A Closure-Preserving Formula-based Extension for Statistical Modeling in SQL"

author:
  - name: "Koki Tsuyuzaki"
    affiliation: "Chiba University`\\\\`{=latex} and RIKEN"
    affiliation2: "Data Science Core, Chiba University`\\\\`{=latex} Chiba, Japan`\\\\[6pt]`{=latex} Omics AI Research Team, Advanced General Intelligence for Science Program (AGIS), RIKEN TRIP Headquarters`\\\\`{=latex} Wako, Japan"
    address: "`%`{=latex}"
    email: "koki.tsuyuzaki@gmail.com"
  - name: "Kentaro Sakamaki"
    affiliation: "Juntendo University`\\\\`{=latex} and Chiba University"
    affiliation2: "Faculty of Health Data Science, Juntendo University`\\\\`{=latex} Tokyo, Japan`\\\\[6pt]`{=latex} Center for Next Generation of Community Health, Chiba University`\\\\`{=latex} Chiba, Japan"
    address: "`%`{=latex}"
    email: "kentaro.sakamaki@gmail.com"
  - name: "Hiromu Nishiuchi"
    affiliation: "Chiba University`\\\\`{=latex} and Chuo University"
    affiliation2: "Data Science Core, Chiba University`\\\\`{=latex} Chiba, Japan`\\\\[6pt]`{=latex} Graduate School of Science and Engineering, Chuo University`\\\\`{=latex} Tokyo, Japan"
    address: "`%`{=latex}"
    email: "hiromunishiuchi@gmail.com"
## Author list, order, affiliations, and corresponding author (first
## author) follow the fbrglm paper. The submission build takes authors
## from paper/vldb/vldbj-template.tex — keep both in sync.

abstract: >
  Statistical modeling is increasingly performed inside SQL database
  systems, where the data already resides, yet existing in-database
  machine learning systems concentrate on computation — scalability,
  algorithm coverage, deployment — and answer a language question only
  implicitly: what should statistical modeling look like *as SQL*? This
  paper proposes FbSQL (Formula-based SQL), a statistical modeling
  domain-specific language that treats SQL's design principles — set
  orientation, declarativeness, closure, order independence, and NULL
  semantics — as hard constraints. Its core design question, which we
  call the *Minimum Atomic Relation* question, asks: for a class of
  statistical models, what is the smallest relation that supports both
  the interpretation of a fitted model and the reconstruction of its
  predictions? For generalized linear models — the proof of concept —
  the answer is a relation of coefficient rows with model-level columns
  and a queryable metadata column that makes it self-contained: models
  are specified with R's formula notation, fitting and prediction consume
  and return relations, and no model object is ever exposed. Because the
  model is data, the language is strictly separated from its execution
  engine — the only thing that crosses the boundary is the model relation
  itself. The open-source PostgreSQL reference implementation
  demonstrates this separation by predicting without R while fitting
  delegates to it, reproducing R's `glm()` and `predict.glm()` at the
  tested precision, and reproducible comparisons against Apache MADlib,
  PostgresML, and Spark MLlib trace observable behavioral differences to
  interface and representation choices.

## Keywords (VLDBJ allows 4-6; used by the submission build, maintained
## in paper/vldb/vldbj-template.tex — keep in sync):
## SQL, PostgreSQL, statistical modeling, formula interface,
## domain-specific language, closure

output:
  html_document:
    toc: true
    number_sections: true
    fig_caption: true
bibliography: references.bib
---

# Introduction

SQL's longevity rests on a small number of design commitments inherited
from the relational model: operations on whole relations, declarative
specification, and — above all — closure, the guarantee that every
operator consumes relations and produces a relation, which is what makes
queries composable into subqueries, views, and pipelines
[@codd1970relational]. These commitments have been articulated, defended,
and fought over for five decades [@date2004introduction; @darwen1995third].
Meanwhile, statistical modeling has moved to where the data lives. Since
the in-database analytics movement took shape [@cohen2009madskills], a
substantial ecosystem has grown: Apache MADlib brings statistical methods
into PostgreSQL and Greenplum as SQL-callable functions
[@hellerstein2012madlib]; PostgresML packages a modern machine learning
stack as a PostgreSQL extension [@postgresml]; Spark MLlib attaches
machine learning to an engine whose query language includes SQL
[@meng2016mllib]; Apache Hivemall embedded training and prediction
directly into HiveQL [@hivemall]; H2O-3 connects a dedicated machine
learning cluster to Spark data [@h2o]; and the commercial success of
managed offerings such as BigQuery ML [@bigqueryml] confirms the demand
from the SQL side.

These systems concentrate, by design, on computation. MADlib and Spark
MLlib emphasize scalability — statistical algorithms that run on parallel
and distributed data. H2O-3 emphasizes algorithm coverage and predictive
performance. PostgresML emphasizes deployment: training, versioning, and
serving models inside the operational database. These are legitimate and
different design objectives, and the systems serve them well; nothing in
this paper argues otherwise.

What has received far less attention is the *language* question. SQL is not
merely an invocation surface; it is a language with a coherent design
inherited from the relational model [@codd1970relational]: operations are
set-oriented, specifications are declarative, every operator consumes and
produces relations, results do not depend on row or argument order, and
NULL has defined semantics. When statistical modeling enters this language,
a series of design questions follows. How should a statistical model be
*specified* — is there a place in SQL for something like the formula
notation that has served statistical practice for three decades
[@chambers1992statistical]? What should a fitting function consume and,
above all, *return* — and if the answer is not a relation, what happens to
the composability that closure provides? Where does the fitted model
*live* — as data the user can query, or as an object, a file, or catalog
state on the side? And what is the shape of prediction — a scalar function
applied row by row, or a mapping from relations to relations? Existing
systems answer these questions implicitly, each making different trade-offs
shaped by its computational objectives — model objects here, procedural
interfaces there, host-language APIs elsewhere. Those answers carry a cost
that is easy to state in the relational model's own terms: the moment the
fitted model is not a relation, closure breaks at the model boundary, and
what composability normally buys is lost exactly where analysis needs it.
A model that is data can be joined against last year's model to see what
changed, audited with a `SELECT` to answer what data it was trained on,
versioned and diffed like any table, and handed to prediction — or to any
future consumer — with no export step and no registry; a model that is an
object, a file, or catalog state can be handled only by its own system's
dedicated functions. Yet the question "what should statistical modeling
look like *as SQL*?" has rarely been posed as a language-design problem in
its own right. That is the problem this paper takes up. We propose FbSQL,
a formula-based statistical modeling DSL that treats SQL's design
principles as hard constraints — most centrally closure: fitting and
prediction consume relations and return relations, and no model object is
ever exposed. The design turns on one question, which we call the
*Minimum Atomic Relation* question: for a given class of statistical
models, what is the smallest relation that supports both the
interpretation of a fitted model and the reconstruction of its
predictions? Generalized linear models serve as the proof of concept; the
claim of the paper is the language design, not the estimator — calling
R's `glm()` from PostgreSQL has long been possible [@plr] and is not, by
itself, a contribution.

This paper makes four contributions. First, it poses the Minimum Atomic
Relation question and answers it for GLMs: a term-grain coefficient
relation with repeated model-level columns and a queryable metadata column
that makes the relation a self-contained model representation. Second, it
distills five design principles — set orientation, declarativeness,
closure, order independence, and NULL semantics — into explicit
constraints for statistical modeling in SQL, and uses them as a lens for
reviewing existing systems. Third, it provides a reference implementation
as an open-source PostgreSQL extension whose numerical behavior is locked
to R's `glm()` and `predict.glm()` by a conformance suite, and whose
predictor runs without R — direct evidence that the model relation, not
engine state, carries the model across the language–engine boundary.
Fourth, it demonstrates the design on a running example — a customer-churn
workflow written entirely in SQL — and evaluates it against five
open-source SQL-oriented machine learning systems, with all comparison
material reproducible from a companion repository. The remainder
of the paper follows this order: related work, the design principles, the
language design, the running example, the reference implementation, the
comparative evaluation, and a discussion of trade-offs and generalization
beyond GLMs.

# Related Work

Systems that bring statistical or machine learning computation to SQL users
differ less in raw capability than in which properties of SQL they choose
to preserve. This section reviews five open-source systems through the lens
of the five design principles introduced above and developed in the next
section, asking of each
system what it emphasizes and what it trades away — not which system is
better. The review rests on two kinds of evidence, and we keep them
distinct throughout. For Apache MADlib (1.21.0 on PostgreSQL 11),
PostgresML (2.7.12 on its bundled PostgreSQL 15), and Spark MLlib (3.5.1),
the observations are *measured*: we reproduced this paper's running example
— a customer-churn logistic regression — on each system inside pinned
Docker environments in the companion repository `FbSQL-experiments`. For
Apache Hivemall and H2O-3 with Sparkling Water, the review is *literature
based*, drawn from the official documentation and recorded claim by claim
with sources in the companion repository; these systems were not executed,
and we do not report anything their documentation does not state. A full
comparison across the design dimensions is maintained as
version-controlled data (`data/related_work.csv`) in `FbSQL-experiments`
and rendered from it as Table 1; the present section confines itself to
the design trade-offs and delegates the dimension-by-dimension detail to
that table.

```{r tab1, echo=FALSE, results='asis'}
## Drop the generator's "%" provenance comments: pandoc does not treat
## them as raw LaTeX and would typeset them as visible text.
tab <- readLines(if (knitr::is_latex_output())
    "tables/related_work.tex" else "tables/related_work.md")
cat(tab[!grepl("^%", tab)], sep = "\n")
```

## A taxonomy of SQL-oriented machine learning systems

The systems fall into three groups according to where the modeling
computation lives relative to the SQL language.

**In-database SQL-ML.** Apache MADlib [@hellerstein2012madlib] and
PostgresML [@postgresml] are PostgreSQL extensions: training and prediction
are invoked as SQL function calls, and the computation runs inside the
database server. These systems share FbSQL's host environment and
invocation surface, which makes the differences in language design the most
directly comparable.

**SQL-on-engine ML.** Apache Spark MLlib [@meng2016mllib] and Apache
Hivemall [@hivemall] attach machine learning to distributed query engines
whose query language is, or includes, SQL. In Spark, SQL defines the input
relations (`spark.sql`) while the modeling itself lives in the
DataFrame/Pipeline API beside the SQL dialect; in Hivemall, training and
prediction are user-defined functions written into HiveQL queries, so the
user never leaves SQL, although — as discussed below — much of the
statistical semantics is left to the query author. Hivemall was retired
from the Apache Incubator in September 2022; we review it nonetheless
because its design occupies a distinctive point that no active system
covers.

**SQL-adjacent ML.** H2O-3 with Sparkling Water [@h2o] is a separate
machine learning cluster with its own frame data structure. Sparkling Water
bridges Spark to it, so data defined by Spark SQL can reach H2O — but only
by conversion from a DataFrame into an H2OFrame, after which the modeling
API is R, Python, or Flow rather than SQL. We include it as the boundary
case: a system that SQL users can reach, built without SQL as a design
target.

BigQuery ML [@bigqueryml] also offers model training through SQL
statements, but it is a proprietary managed service rather than open-source
software: its environment cannot be pinned, versioned, and re-executed the
way the systems above can, so we exclude it from the comparison and return
to it only briefly in the *Discussion* section.

## Design trade-offs

**Apache MADlib** emphasizes in-database scale with full statistical
output (five GLM families with link options; standard errors, *z*
statistics, *p* values), but not specification: predictors are supplied as
an SQL array expression string with user-side one-hot encoding, training
writes two side-effect tables whose coefficient arrays carry no term
names, and prediction is a per-row scalar function. As shown in the
experimental evaluation, these choices surface at the edges of the
workflow, most visibly for factor levels unseen at training time.

**PostgresML** shares FbSQL's deployment surface — a PostgreSQL extension
called through SQL functions — but pursues model lifecycle management:
specification reduces to a task and an algorithm name with no family/link
concept or inferential reporting, the fitted model is a serialized binary
in catalog tables, and prediction resolves through the currently deployed
model of a named project rather than through a value the caller passes.
Its default treatment of text and NULL features likewise departs from
statistical convention, as shown in the experimental evaluation.

**Spark MLlib** demonstrates formula-shaped specification at scale through
its `RFormula` transformer, though with semantics that differ from R's
(frequency-ordered factor levels — a hazard when porting R workflows), and
it preserves closure at the DataFrame level rather than the SQL level:
`transform()` maps DataFrame to DataFrame while the fitted model remains a
JVM object. NULL and unseen-level handling share a single `handleInvalid`
switch with no keep-the-row-with-NULL option; the experimental evaluation
traces both differences on the running example.

**Apache Hivemall** (literature based) makes the opposite trade-off:
everything stays inside HiveQL and the trained model is a genuine
`(feature, weight)` table — the closest any reviewed system comes to
relational closure — but the statistical semantics are left entirely to
the query author: hand-assembled `index:value` feature arrays,
per-algorithm training functions with option-string hyperparameters, and
hand-written prediction SQL over a feature index that carries no term
semantics.

**H2O-3 with Sparkling Water** (literature based) is statistically the
richest GLM reviewed — ten families, links, offsets, weights, and
interactions — but it is not a SQL DSL and does not aim to be one: models
are specified as x/y column lists, live as binary objects in the H2O
cluster, and are reached from Spark SQL only by converting the data out of
the relational world. Its documented defaults (mean imputation,
regularized fits, silent replacement of unseen levels) answer coherently
to predictive-ML conventions rather than to SQL's or R's.

Across the five systems a pattern emerges: each preserves some of the five
principles and relaxes others, and in every case the relaxation follows
from the system's objective — distributed scale for MADlib, lifecycle
operations for PostgresML, pipeline composition for Spark, staying inside
HiveQL for Hivemall, and statistical breadth for H2O. Notably, closure
appears twice in partial form — at the DataFrame level in Spark and in the
model-as-table of Hivemall — but in no reviewed system do set-oriented
invocation, declarative formula semantics, and relation-in / relation-out
closure hold simultaneously at the SQL level.

## Positioning of FbSQL

FbSQL differs from these systems in its design objective rather than in
capability. It asks what the conjunction looks like: SQL functions inside a
DBMS (as in MADlib and PostgresML), model specification with R's formula
semantics (which Spark approximates in shape but not in semantics), a
fitted model that *is* a relation consumed and produced by ordinary queries
(which Hivemall approaches mechanically but without term semantics), and
model metadata that travels inside that relation rather than residing in
system catalogs or host-language objects. Each element exists somewhere in
the reviewed systems; to our knowledge, no system offers their combination
under the five principles, and the running example of this paper is
designed to make the resulting differences observable rather than
rhetorical. The price of this objective is scope: FbSQL's proof of concept
covers two families and deliberately omits offsets, weights, and
interactions — exclusions we return to in the *Discussion* section. The
systems reviewed here are not deficient versions of FbSQL, and FbSQL is not
a replacement for any of them; they optimize different points in a design
space that this paper tries to make explicit.

# Design Principles

FbSQL starts from a deliberately narrow question: if statistical modeling
is to be written *in* SQL, rather than merely invoked *from* SQL, which
properties of the language must be preserved? The contribution is therefore
not the ability to call R's `glm()` from a database — PL/R [@plr] has
provided embedded R in PostgreSQL for many years, and FbSQL indeed uses it
internally — but a language design in which fitting and prediction obey the
same principles as every other SQL operation. This section states the five
principles that FbSQL adopts as design constraints. For each one we describe
its meaning in SQL, the tension it creates with conventional statistical
modeling interfaces, the FbSQL specification decision that resolves that
tension, and the different trade-offs made by existing SQL-oriented machine
learning systems. Throughout, `fit_glm()` and `predict_glm()` serve as the
proof of concept: none of the principles is specific to generalized linear
models, and the *Discussion* section sketches how the same constraints
would shape future functions such as `fit_rf()`, `fit_lgbm()`, and
`fit_xgb()`.

## Set orientation

SQL operates on relations as wholes: a query is a mapping between sets of
rows, and the language provides no construct for visiting rows one at a
time [@codd1970relational]. Statistical modeling interfaces are typically
object-centric instead — fitting produces a model value, and prediction is
a method invoked on that value, with the iteration over observations hidden
in the host language. When such interfaces are transplanted into SQL, the
row-at-a-time perspective tends to resurface on the prediction side: Apache
MADlib [@hellerstein2012madlib] and PostgresML [@postgresml] both expose
prediction as scalar functions evaluated once per row over a hand-supplied
feature array, so the user, not the system, is responsible for assembling
each observation into the form the model expects.

FbSQL keeps both halves of the modeling workflow set-oriented. `fit_glm()`
consumes a relation — any SQL query — and returns its result as a relation
of coefficient rows. `predict_glm()` consumes two relations, the model
relation and the scoring relation, and returns a relation at the granularity
of the scored data. There is no per-row prediction call and no user-managed
feature vector; the unit of every operation is a relation.

## Declarative specification

SQL describes the result to be computed, not the procedure that computes
it. Model specification in existing SQL-oriented systems is frequently
operational — array expression strings in MADlib, hand-assembled
`index:value` feature arrays in Hivemall [@hivemall] — leaving encoding,
interaction columns, and column order as user responsibilities.

FbSQL adopts the formula notation of Chambers and Hastie
[@chambers1992statistical] as its model
specification language: `'churn_flag ~ age + gender'` names a statistical
relationship, and the system derives everything operational from it —
factor levels, treatment contrasts, and the construction of the design
matrix — following the same conventions as R's `glm()`. Spark MLlib
[@meng2016mllib] demonstrates that formula-shaped specification scales,
though with semantics that differ from R's; PostgresML reduces
specification to a task and an algorithm name — concise, but the
statistical model itself is no longer something the user states.

## Closure: relation-in / relation-out

Relational algebra is closed: every operator consumes relations and
produces a relation, which is what makes queries composable through
subqueries, views, and common table expressions [@codd1970relational].
This is where statistical modeling diverges most sharply from SQL
practice. In nearly every system the fitted model is an opaque value that
lives outside the data model: an S object in R, a `PipelineModel` in Spark
(a JVM object graph persisted as a directory), a serialized binary model in
PostgresML's catalog tables, or, in MADlib, a pair of side-effect output
tables written by a void-returning training procedure. Once the model is
not a relation, composition stops at the model boundary: the fitted object
can only be handled by the system's own dedicated functions.

The central design decision of FbSQL is that **the fitted model is a
relation** — one row per design matrix column, with model-level columns
alongside, specified in full in the *Language Design* section. The model
can be materialized, queried, joined, and audited with ordinary SQL, and
`predict_glm()` closes the loop by consuming that relation together with a
scoring relation. That the relation is a complete representation of the
model, rather than a projection of hidden state, is demonstrated by the
reference implementation itself: its predictor contains no R and computes
purely from the public relation.

Keeping the output to exactly one relation costs two deliberate
denormalizations — model-level scalars and the `metadata` JSONB column are
repeated on every row — a trade-off the *Language Design* section
specifies and the *Discussion* section weighs. Unlike a binary model
object, the JSONB metadata remains inspectable from SQL. Existing systems
occupy different points on this spectrum rather than simply lacking the
property — closure at the DataFrame level in Spark, a model table without
term semantics in Hivemall, procedural interfaces in MADlib and PostgresML
— as reviewed in the previous section.

## Order independence

Relations are unordered, and a query result carries no row order unless an
`ORDER BY` is written. This principle has two consequences for FbSQL. The
first concerns row order: FbSQL guarantees nothing about the row order of
the model relation, so `predict_glm()` matches coefficients to terms
by name and cross-checks the set of terms against the `coef_terms` metadata
field, never by position — the contrast is MADlib's parallel coefficient
arrays, where position is the only semantics.
The second consequence concerns argument order. Statistical functions
accumulate arguments quickly (`relation`, `formula`, `family`, and in the
future offsets, weights, or tuning parameters), and positional calls become
fragile as the list grows. FbSQL therefore specifies its calls with named
notation — `fit_glm(relation => ..., formula => ..., family => ...)` — so
that invocations are order independent and self-documenting. Named notation
is PostgreSQL (and Oracle) syntax rather than standard SQL; we return to
this limitation in the *Discussion* section.

## NULL semantics

SQL treats NULL as *unknown* under a three-valued logic, and NULLs
propagate through expressions. The statistical counterpart is missing
data, and existing systems answer independently of SQL's convention — hard
errors (PostgresML), silently dropped rows (Spark's
`handleInvalid = "skip"`), default mean imputation (H2O-3 [@h2o]) —
choices defensible in context, but none following from the semantics of
the surrounding query language. Relational theory itself carries a
long-standing critique of NULL and three-valued logic
[@date2004introduction; @darwen1995third]; FbSQL takes no side in that
debate. Its constraint is fidelity to the language as practiced, and NULL
is part of SQL as practiced.

FbSQL aligns the two traditions where they already agree. At fitting time
it performs complete-case analysis, exactly matching the `na.omit` default
of R's `glm()`, and reports rather than hides the effect through three
model-level columns (`n_obs`, `n_used`, `n_dropped`). At prediction time a
NULL predictor propagates to a NULL prediction — the result that SQL
expression evaluation of the linear predictor would itself produce, and the
relational rendering of R's `NA` prediction. No value is imputed, no row is
dropped, and no error interrupts the query: the output keeps the
granularity of the input, and NULL means exactly what it means everywhere
else in SQL.

## From glm to a family of modeling functions

The five principles, not the estimator, define FbSQL: `glm()` was chosen
as the proof of concept because it is small enough to specify completely
and rich enough to exercise every principle. For any future `fit_xxx()`,
each principle poses the same design question, culminating in the *Minimum
Atomic Relation* question that the next section states precisely and the
*Discussion* section extends to tree ensembles. The claim of this paper is
the language design; the estimator behind `fit_glm()` is deliberately
ordinary.

# FbSQL Language Design

This section specifies FbSQL as a language: what the user writes and what
the system guarantees. How the reference implementation computes it —
PL/R, testing, packaging — is deferred to the implementation section; here
we describe only what was designed.

## Design goals

Four goals follow directly from the design principles. First, statistical
modeling should be *written as* SQL rather than *called from* SQL: fitting
and prediction are ordinary functions usable inside queries, views, and
`CREATE TABLE AS`, not commands that leave the query language. Second, the
formula is the first-class model-specification construct: the user states
the statistical relationship, and the language defines everything
operational that follows from it. Third, every modeling operation is
relation-in / relation-out. Fourth — the contrapositive of the third — no
model object is ever exposed: the fitted model is data, not a handle. The
proof of concept binds these goals to two functions:

```sql
fbsql.fit_glm(
    relation text,   -- SQL string: training relation
    formula  text,   -- R formula
    family   text DEFAULT 'gaussian'
) RETURNS SETOF fbsql.glm_fit

fbsql.predict_glm(
    relation text,   -- SQL string: relation to score
    model    text,   -- SQL string: a model relation
    on_new_levels text DEFAULT 'error'  -- | 'na'
) RETURNS SETOF record
```

Both functions live in the `fbsql` schema; the examples in this paper
assume `SET search_path TO fbsql, public;` and write them unqualified.
Figure 1 gives the overview: every arrow in it is a relation.

```{r fig1, echo=FALSE, out.width="100%", fig.env="figure*", fig.cap="FbSQL system overview. The upper layer is the language: fitting maps a training relation and a formula to a model relation, and prediction maps the model relation and a scoring relation to a prediction relation --- every arrow is a relation. The model relation is the central artifact: queryable, joinable, auditable, and self-contained through its metadata column. The engine below the boundary is replaceable without changing any SQL above it."}
knitr::include_graphics(if (knitr::is_latex_output())
    "figures/figure1_system_overview.pdf" else "figures/figure1_system_overview.svg")
```

## Formula-based model specification

Fitting the running example's churn model is a single call, shown here at
signature level and in full in the *Running Example* section:

```sql
fit_glm(
    relation => $$ SELECT ... FROM customer ... $$,
    formula  => 'churn_flag ~ age + gender',
    family   => 'binomial')
```

The three arguments divide the specification cleanly. `relation` is a SQL
string — dollar-quoted by convention — defining the training data. Because
it may be any query, the modeling layer introduces no data-access
sublanguage of its own: filtering, joining, and feature construction happen
in ordinary SQL before the model ever sees the data, and anything SQL can
express can be trained on. `formula` carries the statistical specification
in the notation of Chambers and Hastie [@chambers1992statistical]. We
adopt it rather than invent
a new syntax for three reasons: it has carried three decades of
statistical-modeling practice, it is compact and declarative — it names a
relationship, not a procedure — and it comes with precise, widely
understood semantics. FbSQL fixes those semantics to be exactly those of
R's `glm()` [@rcore]: character columns become factors with sorted levels,
the first level is the reference under treatment contrasts, and rows with
missing values are dropped as in `na.omit`. Fixing the semantics by
reference to an existing system makes the specification *testable*:
conformance of any implementation can be checked against R output, and the
reference implementation's conformance suite does exactly that. The formula
is therefore not merely a string argument; it is the model-specification
sublanguage of the DSL, and the derivation of the design matrix — encoding,
contrasts, reference levels — is defined by the language rather than
performed by the user. `family` names the error distribution, with the
canonical link implied in the proof of concept (gaussian/identity,
binomial/logit).

The call style is part of the specification: named-argument invocation
(`relation => ...`, `formula => ...`) is the normative form, applying the
order-independence principle to arguments as the list grows. The notation
is PostgreSQL and Oracle syntax rather than standard SQL, a point we
return to in the *Discussion* section.

## Relation representation of statistical models

The central design decision of FbSQL is what `fit_glm()` returns. It does
not return a handle to a model; it returns the model itself, as a relation
of type `fbsql.glm_fit` with seventeen columns at two granularities:

```{r spec-table, echo=FALSE, results='asis'}
## The column specification, per output format: markdown's pipe table
## becomes a longtable in LaTeX, which twocolumn svjour3 cannot host, so
## the LaTeX build gets a plain non-floating tabular instead.
spec <- list(
  c("\\texttt{term}", "term", "design matrix column name, as in R's coefficient table"),
  c("\\texttt{estimate}", "term", "coefficient"),
  c("\\texttt{std\\_error}", "term", "standard error"),
  c("\\texttt{statistic}", "term", "\\emph{t} value (gaussian) / \\emph{z} value (binomial)"),
  c("\\texttt{p\\_value}", "term", "\\emph{p} value"),
  c("\\texttt{conf\\_low\\_95}, \\texttt{conf\\_high\\_95}", "term", "Wald 95\\% confidence bounds"),
  c("\\texttt{family}, \\texttt{link}, \\texttt{formula}", "model", "the specification, echoed back"),
  c("\\texttt{n\\_obs}, \\texttt{n\\_used}, \\texttt{n\\_dropped}", "model", "complete-case accounting"),
  c("\\texttt{aic}, \\texttt{deviance}, \\texttt{null\\_deviance}", "model", "model-level fit measures"),
  c("\\texttt{metadata}", "model", "JSONB; everything prediction needs (below)"))
if (knitr::is_latex_output()) {
  cat("\\begin{center}\\footnotesize\n",
      "\\begin{tabular}{@{}p{2.6cm}p{0.8cm}p{4.2cm}@{}}\n\\toprule\n",
      "Column & Grain & Meaning \\\\\n\\midrule\n", sep = "")
  for (r in spec) cat(paste(r, collapse = " & "), "\\\\\n", sep = "")
  cat("\\bottomrule\n\\end{tabular}\n\\end{center}\n")
} else {
  md <- gsub("\\\\([_%])", "\\1", vapply(spec, function(r)
      paste0("| ", gsub("\\\\texttt\\{([^}]*)\\}", "`\\1`",
             gsub("\\\\emph\\{([^}]*)\\}", "*\\1*", paste(r, collapse = " | "))), " |"),
      character(1)))
  cat("| Column | Grain | Meaning |", "|---|---|---|", md, sep = "\n")
}
```

One row corresponds to one design matrix column — `(Intercept)`, each
numeric predictor, each non-reference factor level — so the churn model of
Figure 2 is a four-row relation with terms `(Intercept)`, `age`,
`genderM`, and `genderOther`. The term-level column names follow the
vocabulary R users know from model-tidying conventions (`term`,
`estimate`, `std_error`, `statistic`, `p_value`), adjusted to legal SQL
identifiers, so the relation is immediately readable from both sides of
the language boundary.

```{r fig2, echo=FALSE, out.width="100%", fig.env="figure*", fig.cap="The running example as one relational circuit. The same customer table supplies the 2025 cohort for fitting and the 2026 cohort for scoring, and the fitted churn model passes between the two functions as an ordinary relation --- nothing outside the diagram carries state. The awkward rows of production scoring data (a NULL age, an unseen gender level) are handled inside SQL's own semantics rather than outside the relational world."}
knitr::include_graphics(if (knitr::is_latex_output())
    "figures/figure2_running_example.pdf" else "figures/figure2_running_example.svg")
```

Returning a relation rather than an object has three consequences that
motivate the design. First, *composition*: the model can be materialized
(`CREATE TABLE AS`), stored, joined, and filtered like any other data —
two refits can be compared by joining their relations on `term`, and a
model-level quantity is one query away (`SELECT DISTINCT aic FROM
logit_model`). Second, *inspectability*: there is nothing about the fitted
model that `SELECT` cannot show, which contrasts with the serialized or
host-language model representations reviewed in the *Related Work*
section. Third, *implementation independence*: because the language defines
the model as data with specified columns, any implementation — in any
language — that produces and consumes this relation is a conforming one;
the model representation is part of the language, not of the engine.

Two granularities coexist in one relation, and this is deliberate.
Model-level values (`aic`, `n_used`, `family`, ...) are repeated on every
term row — a denormalization, since these columns depend on the model, not
the term. The normalized alternative, splitting the output into coefficient
and model-summary relations, was rejected because a multi-relation return
gives up closure — the property the design exists to preserve. FbSQL
accepts the redundancy and documents it as the explicit price of
one-relation-out; the *Discussion* section weighs this trade-off.

The same reasoning shapes the seventeenth column. A coefficient table alone
is not a complete model representation: reconstructing predictions requires
the factor levels seen at training time, the contrast scheme, the term
structure, and the column types — information that R keeps inside its
model object. FbSQL stores it in a `metadata` JSONB column, repeated on
every row like the other model-level columns. For a model
`y ~ x1 + gender` fitted on a numeric predictor and a three-level factor,
the value is (field for field as produced by the reference
implementation; reflowed for column width):

```json
{
  "meta_version": 1,
  "response": "y",
  "term_labels": ["x1", "gender"],
  "intercept": true,
  "data_classes": {"y": "numeric",
                   "x1": "numeric",
                   "gender": "factor"},
  "xlevels": {"gender": ["F", "M", "Other"]},
  "contrasts": {"gender": "contr.treatment"},
  "coef_terms": ["(Intercept)", "x1",
                 "genderM", "genderOther"]
}
```

The schema is versioned (`meta_version`) because the field layout is part
of the language's API, and it never duplicates the flat columns (`family`,
`formula`, the counters), keeping a single source for each fact. The one
deliberate redundancy is `coef_terms`, the canonical set of coefficient
terms: since row order is never guaranteed, a consumer matches coefficients
by name and uses `coef_terms` to detect missing or duplicated rows.
`xlevels` answers what no coefficient table can — the reference level `F`
appears in no `term` — and enables the unseen-level checks of the next
subsection. Storing this as JSONB is itself a trade-off: it is a departure
from flat relational columns, but it keeps the output to one relation, and
unlike an opaque model object it remains queryable — `metadata ->
'xlevels'` is an ordinary SQL expression. With `metadata`, the relation is
a *self-contained* model representation: nothing needed for prediction
lives outside it.

## Prediction interface

Prediction consumes two relations and returns one, again at signature
level (the full query appears in the *Running Example* section):

```sql
predict_glm(
    relation => $$ SELECT ... FROM customer ... $$,
    model    => $$ SELECT * FROM logit_model $$
) AS p(customer_id varchar, age integer,
       gender varchar,
       churn_flag_predicted double precision)
```

Both inputs are relations on equal footing. The scoring relation is given
as a SQL string as in `fit_glm()`; the model is *also* a SQL string — any
query returning a model relation (of type `glm_fit`) qualifies, whether a
materialized
table, a view, or a filtered copy. The model is passed as data, not as a
name resolved against system state: there is no registry, no deployment
indirection, and no session dependence — scoring is a pure function of the
two relations. The output is the scoring relation's rows with one appended
column, `<response>_predicted`, named from the metadata (here
`churn_flag_predicted`, a probability on the response scale). The output
keeps the granularity of the input — one row in, one row out — and carries
no row-order guarantee, per the order-independence principle; coefficients
are matched to terms by name and validated against `coef_terms`, never by
position. One ergonomic cost is visible in the example: because the output
schema is "the input columns plus one derived column", it cannot be
declared statically in PostgreSQL's type system, so `predict_glm()` returns
`SETOF record` and the caller attaches a column definition list. We record
this as a limitation of the current specification and discuss alternatives
in the *Discussion* section.

Missing and unseen values follow the NULL-semantics principle. A NULL
predictor yields a NULL prediction — the row is kept, and NULL means
unknown. Factor levels not seen at training time are governed by
`on_new_levels`, an argument of `predict_glm()` rather than a property of
the model: whether an unknown level should abort the query or produce an
unknown prediction is a decision of the *consumer* of the predictions, not
a fact about the fitted model, so it does not belong in `metadata`. The
default `'error'` fails loudly, naming the offending level and the known
levels from `xlevels`; `'na'` keeps the affected rows with NULL
predictions, preserving output granularity. This design — detection from
stored levels, an explicit caller-side policy, error as the default — is
carried over from fbrglm [@fbrglm], an R package by the first author whose
formula-based interface confronted the same unseen-level problem; FbSQL
restates that answer in relational terms.

## Generalization beyond GLM

Nothing in the interface above is specific to generalized linear models.
The pattern is a pair of conventions — `fit_xxx(relation, formula, ...)`
returning a model relation, and `predict_xxx(relation, model)` returning
the scored relation — plus one design question that each new estimator
family must answer, which we call the *Minimum Atomic Relation* question:

> What is the smallest relation that supports both the interpretation of a
> fitted model and the reconstruction of its predictions?

For GLMs the answer is the seventeen-column relation above: term-grain
coefficient rows, repeated model-level columns, and the metadata needed to
rebuild the design matrix. For tree ensembles — future `fit_rf()`,
`fit_lgbm()`, `fit_xgb()` — the natural grain is not the coefficient but
the split: a model relation whose rows are nodes (tree, node, split
variable, threshold, leaf value), with ensemble-level columns and metadata
playing the same roles as here. The interface conventions — named
arguments, relation-valued model input, `<response>_predicted`
output, caller-side policies for unseen data — carry over unchanged. The
*Discussion* section sketches these designs; the point of this section is
that the language, not the estimator, is the fixed part of FbSQL.

# Running Example

This section walks through FbSQL from the user's side: a complete
fit-and-predict workflow written in nothing but SQL. The example is not
illustrative pseudocode — it ships verbatim as a conformance test in the
reference implementation (`test/sql/running_example.sql`), and every number
shown below is locked by that suite and verified against R's `glm()` and
`predict.glm()` to four decimals in the companion repository. Figure 2
summarizes the flow: the `customer` table feeds `fit_glm()`, which yields a
model relation; the model relation and the scoring rows feed
`predict_glm()`, which yields a prediction relation.

<!-- Figure 2 itself is placed at its first reference in the Language
     Design section; this section references it again in prose. -->

## Dataset

The example uses a single `customer` table (Table 2) with five columns:
`customer_id` (`VARCHAR`), the customer identifier; `created_at`
(`TIMESTAMP`), when the customer signed up; `age` (`INTEGER`) and `gender`
(`VARCHAR`, taking the values `F`, `M`, and `Other` in the training data),
the predictors; and `churn_flag` (`BOOLEAN`), whether the customer churned
— the response. Twelve customers created in 2025 have observed churn
outcomes and serve as training data. Five customers created in 2026 are to
be scored; their `churn_flag` is NULL because the outcome is not yet known.
The 2026 rows deliberately include two realities of production scoring
data: customer `c104` has a NULL `age`, and customer `c105` reports a
gender (`Nonbinary`) that does not occur in the 2025 training data. The
scenario is the everyday one: learn from last year's customers, score this
year's.

```{r tab2, echo=FALSE, results='asis'}
tab <- readLines(if (knitr::is_latex_output())
    "tables/customer_dataset.tex" else "tables/customer_dataset.md")
cat(tab[!grepl("^%", tab)], sep = "\n")
```

## Model fitting

Fitting the churn model is one SQL statement:

```sql
CREATE TEMPORARY TABLE logit_model AS
SELECT *
FROM fit_glm(
    relation => $$
        SELECT churn_flag, age, gender
        FROM customer
        WHERE DATE_PART('YEAR', created_at) = 2025
    $$,
    formula => 'churn_flag ~ age + gender',
    family  => 'binomial');
```

The three arguments carry the three concerns of the *Language Design*
section. `relation` selects the training data with ordinary SQL — here, the
2025 cohort; any filtering or feature construction would go in this query.
`formula` states the model: churn as a function of age and gender.
`family` picks logistic regression. Everything operational happens without
user involvement: `gender` becomes a factor with reference level `F`, and
the boolean response is accepted directly. The result is a table like any
other, and selecting from it shows the fitted model:

```sql
SELECT term,
       round(estimate::numeric, 4)  AS estimate,
       round(std_error::numeric, 4) AS std_error
FROM logit_model ORDER BY term;
```
```
    term     | estimate | std_error
-------------+----------+-----------
 (Intercept) | -12.1071 |    7.1895
 age         |   0.2981 |    0.1669
 genderM     |  -0.4305 |    2.5525
 genderOther |  -0.7049 |    2.8735
```

One row per term, exactly as R would report it. The model-level columns
repeat on every row — here `family` = binomial, `link` = logit, and
`n_obs` = `n_used` = 12 with `n_dropped` = 0 — and the relation also
carries the `metadata` column of the *Language Design* section, to which
we return below.

## Prediction

Scoring the 2026 customers consumes two relations — the data to score and
the model, both passed as queries — and returns a relation:

```sql
SELECT customer_id,
       round(churn_flag_predicted::numeric, 4)
           AS churn_flag_predicted
FROM predict_glm(
    relation => $$
        SELECT customer_id, age, gender
        FROM customer
        WHERE DATE_PART('YEAR', created_at) = 2026
          AND customer_id <> 'c105'
    $$,
    model => $$ SELECT * FROM logit_model $$
) AS p(customer_id varchar, age integer,
       gender varchar,
       churn_flag_predicted double precision)
ORDER BY customer_id;
```
```
 customer_id | churn_flag_predicted
-------------+----------------------
 c101        |               0.0406
 c102        |               0.9794
 c103        |               0.4280
 c104        |
```

The model argument deserves a second look: `$$ SELECT * FROM logit_model $$`
passes the fitted model *as data*. Any query returning a model relation
would do; there is no model registry to consult and no name to
resolve. The output is the input relation's rows plus one column,
`churn_flag_predicted` — a churn probability, named after the response in
the formula. Customer `c104`, whose `age` is NULL, is neither dropped nor
rejected: the row is kept and its prediction is NULL, which is what NULL
means in SQL — unknown in, unknown out.

Customer `c105` was excluded above because its `gender` value `Nonbinary`
never occurred in the training data, and under the default policy scoring
it stops the query with an explicit error:

```
ERROR:  predict_glm: factor 'gender' has new
level 'Nonbinary' not seen at fit time (known
levels: ["F", "M", "Other"]); use
on_new_levels => 'na' to return NULL for such
rows
```

Passing `on_new_levels => 'na'` instead keeps all five rows and returns a
NULL prediction for `c105` alone. The two policies reflect one design idea:
what to do with an unseen level is the *caller's* decision, not the
model's, so it is an argument of `predict_glm()`. The default fails loudly
— silent number-producing behaviors for unseen categories are among the
differences documented in the *Related Work* comparison — and the `'na'`
alternative preserves the output granularity while marking exactly the
affected rows as unknown.

## Inspecting the fitted model

Because the fitted model is a relation, everything a user might ask of it
is a query, not a method call. Model-level quantities are one `SELECT`
away:

```sql
SELECT DISTINCT aic, deviance, null_deviance
FROM logit_model;
```

The training-time conditions travel with the model in its `metadata`
column and answer questions the coefficient table cannot. Which factor
levels did the model see? —

```sql
SELECT DISTINCT metadata -> 'xlevels' FROM logit_model;
```
```
 {"gender": ["F", "M", "Other"]}
```

The reference level `F` appears in no coefficient row, yet it is right
there in `xlevels`; this is also precisely the information behind the
unseen-level error above. And because the model is a relation, it composes
with SQL itself. Refitting next year's model and joining it against this
year's compares two models term by term, with no export step and no
model-object API:

```sql
SELECT a.term,
       a.estimate AS estimate_2025,
       b.estimate AS estimate_2026
FROM logit_model AS a
JOIN logit_model_2026 AS b USING (term);
```

Nothing in this section serialized an object, called a method, or consulted
system state: the model was created by a query, stored as a table, passed
to prediction as a query, and audited by queries.

## Summary

The workflow was four SQL statements: create the model relation, read it,
score with it, and inspect it. Both `fit_glm()` and `predict_glm()`
consumed relations and produced relations; the model itself was a table
throughout its life; and at no point did the user touch anything but SQL.
Note also what the walkthrough did *not* depend on: no step revealed — or
needed to reveal — what computes behind the two functions. That opacity is
the language boundary working as designed, and it is what makes the
implementation replaceable. The *PostgreSQL Extension Implementation*
section describes the reference implementation on the other side of that
boundary, including the fact that prediction runs without R at all — a
direct consequence of the model relation being, as the *Language Design*
section specified, a self-contained representation.

# PostgreSQL Extension Implementation

This section describes the reference implementation of FbSQL — not how the
code works line by line, but which implementation decisions were made and
why. The guiding theme is that every decision below is *replaceable*: the
language of the previous sections is the fixed part, and the
implementation exists to demonstrate that the specification can be met.

## Reference implementation

FbSQL is implemented as an ordinary PostgreSQL extension: it installs with
`CREATE EXTENSION`, places its two functions in a dedicated schema, and
uses no mechanism outside PostgreSQL's standard extension infrastructure.
Inside the extension, `fit_glm()` is written in PL/R [@plr], the procedural
language that embeds an R interpreter in the database server, and delegates
the actual estimation to R's `stats::glm()` [@rcore].

The reason is conformance, not novelty. The *Language Design* section fixed
FbSQL's formula semantics by reference to R's `glm()`; an
implementation that runs `glm()` itself therefore satisfies those semantics
by construction, and the engineering effort concentrates where the
contribution lies — the relation representation — rather than on
reimplementing iteratively reweighted least squares. We emphasize what this
choice does *not* mean. Embedding R in PostgreSQL is not new — PL/R has
been available for two decades — and FbSQL claims no contribution there.
Nor is PL/R part of the language: no R value, object, or error surface
reaches the user, whose API is SQL alone, and the next subsection makes the
independence explicit. PL/R is simply the fastest honest route to a
reference implementation whose numbers are correct by construction. (One
operational consequence is worth recording: PL/R is an untrusted language,
so installing the extension requires superuser privileges, with execution
then granted to regular users.)

## Separation of language and execution engine

The implementation is organized around a strict separation between the
*language specification* and the *execution engine*. The specification
comprises everything the previous sections defined: the function
signatures and named-argument call style, the formula semantics fixed to
R's `glm()`, the seventeen-column model relation, the versioned metadata
schema, the NULL and unseen-level policies, and the error behavior. The
engine is whatever machinery produces and consumes relations with exactly
those properties — currently an embedded R interpreter behind `fit_glm()`
and plain PL/pgSQL behind `predict_glm()`.

```{r fig3, echo=FALSE, out.width="100%", fig.env="figure*", fig.cap="The implementation as layers. The language specification is the fixed part; the PostgreSQL extension is one realization of it, with R behind fitting and no R behind prediction, so the model relation is the boundary artifact --- the only thing that crosses between the two engines. The verification layer enforces conformance continuously, which is what makes the engines replaceable in practice and not only in principle."}
knitr::include_graphics(if (knitr::is_latex_output())
    "figures/figure3_implementation_layers.pdf" else "figures/figure3_implementation_layers.svg")
```

Figure 3 sketches the layering. The property that makes the separation
real, rather than aspirational, is that the interface between the layers
is *data*: the model relation and its metadata column carry everything
that crosses the boundary, so nothing about the engine leaks into the
language. The estimation behind `fit_glm()` could be replaced by a C
implementation of IRLS, a GPU solver, a call into MADlib's existing
distributed estimators, or a remote engine, and no user query — none of
the SQL in the *Running Example* section — would change. Conformance of
any such replacement is checkable, not arguable: it must produce the same
relation, and the conformance suite described below tests exactly that.
This is also why the reference implementation's use of R is not a
dependency of the design: it is the first engine, not the engine.

## Fitting: generating the model relation

<!-- Heading avoids the literal function name (a habit from the earlier
     JSS build, whose class broke on underscores in headings; harmless to
     keep). -->

`fit_glm()` is a PL/R function that turns one relation into another. It
executes the caller's `relation` query through the server-side query
interface, assembles the result as an R data frame, applies the documented
type conventions deterministically (character columns are converted to
factors explicitly, with sorted levels and treatment contrasts, rather than
left to R's implicit coercions), fits the model with `stats::glm()`, and
then flattens the fitted object into the seventeen-column relation of the
*Language Design* section — coefficient rows from the summary, Wald
confidence bounds (chosen over profile-likelihood intervals for
determinism and cost; the choice is documented as part of the
specification), the complete-case counters, and the model-level measures.
The `metadata` column is built inside the same function from the fitted
object's term, level, and contrast information. Nothing of R survives the
call: the model object is garbage-collected when the function returns, and
the relation is the only artifact. In the terms of this paper, `fit_glm()`
is best read not as "calling R from SQL" but as *generating a relation*
that happens to use R as its oracle.

## Prediction without R

`predict_glm()` contains no R at all. It is implemented in PL/pgSQL,
PostgreSQL's native procedural SQL dialect, and computes predictions
purely from the two relations it receives: it matches coefficient rows to
terms by name, validates them against the metadata's canonical term set,
rebuilds the treatment-contrast columns for factors from the stored
levels, evaluates the linear predictor as a SQL expression over the
scoring relation, and applies the inverse link. This is an implementation
decision with a rhetorical point. If the model relation were not a
complete representation of the fitted model, a second implementation in a
different language, consuming only the public relation, could not
reproduce R's predictions — yet it does, to four decimals, across every
test in the suite. The R-free predictor is thus an existence proof of the
*Language Design* section's central claim: the relation, not any hidden
engine state, is the model. It also prefigures the replacement path
described above — prediction has already crossed the engine boundary
once.

One error-handling decision deserves a paragraph, because it encodes a
lesson about embedding interpreters in database servers. FbSQL validates
its own preconditions eagerly and raises prefixed, explicit errors for a
bad family, an unparsable formula, a missing column, or an empty relation.
But the execution of the caller's `relation` SQL is deliberately *not*
wrapped in R-level error handling. When a query fails inside the server's
query interface, PostgreSQL aborts the surrounding transaction; during
development, catching that failure in R and raising a friendlier error
from the handler crashed the backend, because the R-level error machinery
ran on top of an already-aborted transaction state. The reference
implementation therefore lets native PostgreSQL errors propagate
unchanged. The general lesson: an embedded interpreter must defer to the
host's transaction semantics, and defensive wrapping at the boundary can
be worse than none.

## Verification strategy

The specification is only as credible as its enforcement, so the
implementation invests in verification rather than features. Four layers
are in place. First, the extension carries a `pg_regress` conformance suite
— PostgreSQL's own testing harness — whose eleven test files function as
executable conformance tests of the specification: output columns, factor
conventions, complete-case accounting, metadata content, NULL propagation,
unseen-level policies, and every documented error message. Second, *R
parity*: the same fixtures are defined literally both in the SQL tests and
in a reference R script, and the expected outputs are locked to
`stats::glm()` and `predict.glm()` at four-decimal rounding (rounding
absorbs platform floating-point variation; ordering is pinned to a
locale-independent collation). The specification's "semantics by reference
to R" is thereby a tested property, not a stated intention. Third, the
development environment is pinned with Docker — no standard PostgreSQL
image ships PL/R, so a purpose-built image fixes the PostgreSQL, PL/R, and
R versions — which makes the suite's results reproducible rather than
machine-dependent. Fourth, GitHub Actions runs the full build and
conformance suite on every commit, so conformance is guaranteed
*continuously*: a change that breaks R parity cannot land silently.
Finally, the running example of the previous section is itself one of the
conformance tests, executed verbatim by the suite — every number printed in
this paper's walkthrough is re-verified on every commit, and the same
workflow is reproduced against the comparison systems in the companion
repository, as the next section describes.

# Experimental Evaluation

For a language-design proposal, evaluation does not mean speed, and this
section reports no performance benchmark — computational performance is
explicitly outside FbSQL's claims. The evaluation instead answers three
questions. Does the reference implementation conform to its specification?
Does the proof of concept work end to end? And when the same statistical
task is expressed on existing systems, what does each system's design
require of the user? All materials — scripts, pinned Docker environments,
raw outputs, and summary tables — live in the companion repository
`FbSQL-experiments`, with results committed so that changes surface as
diffs; this section reports what was observed and defers interpretation to
the *Discussion* section.

## Conformance to R

The *Language Design* section fixed FbSQL's formula semantics by reference to R's
`glm()` and `predict.glm()`, so the first evaluation question is
conformance. Independently of the extension's own conformance suite, the
companion repository re-runs the running example and compares every
number against R: the four coefficients, the four standard errors, and
all five predictions — including the two that must be NULL — at
four-decimal rounding. All thirteen comparisons agree
(`results/summary/running_example_parity.csv`). Combined with the
`pg_regress` suite of the previous section, which locks the same property
across all eleven test files on every commit, conformance to the
specification is a continuously checked result, not a stated intention.

## Running example end to end

The second question is whether the proof of concept holds together as a
whole, not merely as parts. The running example exercises the full MVP
surface in one workflow: a binomial fit with a numeric predictor, a
three-level factor, and a boolean response; a model relation whose
`metadata` is then actually consumed — the stored levels and contrasts
drive prediction, not merely document it; scoring that keeps the
NULL-`age` row with a NULL prediction; the default unseen-level policy
stopping the query with the documented error for the `Nonbinary` row; and
the `'na'` policy returning NULL for exactly that row while scoring the
rest. Every statement and every number shown in the *Running Example*
section is executed verbatim by the conformance suite and reproduced in
the companion repository.

## Comparative evaluation

The third question is comparative: the same churn workflow — same
seventeen customers, same model, same scoring set — was reproduced on
Apache MADlib 1.21.0 (on PostgreSQL 11), PostgresML 2.7.12 (on its bundled
PostgreSQL 15), and Spark MLlib 3.5.1 (local mode), each inside a pinned
Docker environment in the companion repository. The two Tier-3 systems of
the *Related Work* section (Hivemall; H2O-3 with Sparkling Water) were
compared from documentation only and are not re-examined here. The
dimension-by-dimension results are in the comparison table introduced in
the *Related Work* section and in per-system design-notes tables in the
companion repository (`results/summary/*_api_design_notes.csv`); this
subsection reports the design differences the reproduction made
observable, not the tables' every cell. Figure 4 recalls the three-group
taxonomy in which the compared systems sit.

```{r fig4, echo=FALSE, out.width="100%", fig.env="figure*", fig.cap="Three ways SQL reaches machine learning, grouped by where the modeling computation lives relative to the SQL language. The groups differ in what the user must leave SQL to do and in where the fitted model resides: relations, side tables, or catalogs inside the database; host-language objects beside the engine's SQL dialect; or an external cluster reached by data conversion. The grouping organizes the comparison; it does not rank the systems."}
knitr::include_graphics(if (knitr::is_latex_output())
    "figures/figure4_system_taxonomy.pdf" else "figures/figure4_system_taxonomy.svg")
```

**MADlib.** Numerically, MADlib fits the same model: all four coefficients
agree with FbSQL and R to four decimals, and one standard error differs in
the fourth decimal (attributable to a different IRLS convergence
tolerance); predictions for the regular scoring rows agree to four
decimals. The differences are therefore almost entirely interface design.
Reproducing the workflow required the user to one-hot encode `gender` by
hand before training, to pass the predictors as an array expression
string, and to score through a per-row scalar function joined against the
model table's coefficient array. Two consequences were observed directly:
the model tables store coefficients as position-indexed arrays without
term names, so the mapping from coefficient to meaning exists only at the
call site; and the unseen-level customer `c105` was scored silently — the
hand-built dummy columns are all zero for an unknown level, so MADlib
returned a probability as if `c105` were the reference gender, with no
error, warning, or NULL.

**PostgresML.** Here numeric agreement is not expected and was not
observed: PostgresML answers a different specification. Its API takes a
task and an algorithm name rather than a formula and a family;
`pgml.predict()` returned class labels (with probabilities available from
a separate function), and no standard errors, *p* values, or confidence
intervals exist to compare. The reproduction surfaced three observable
design facts: text predictors were label-encoded to ordinal reals by
default (`F` = 1, `M` = 2, `Other` = 3), so a linear model receives an
artificial ordering of genders; a NULL feature at prediction time raised a
hard error, so the NULL-`age` row had to be guarded by hand; and the
fitted model exists only as a serialized binary in PostgresML's catalog
tables, reached at scoring time through the project name's currently
deployed model. Although PostgresML shares FbSQL's host — both are
PostgreSQL extensions called from SQL — none of the running example's SQL
shape carries over.

**Spark MLlib.** Spark's generalized linear regression is an IRLS GLM, and
its predictions agree with R and FbSQL to four decimals. The fitted model,
however, is presented differently: `RFormula` ordered the factor levels by
frequency and took the *least frequent* level (`Other`) as the reference,
where R and FbSQL take the first sorted level (`F`), so the two coefficient
tables disagree term by term. The companion repository verifies numerically
that they are the same model under reparameterization — each Spark
coefficient is an exact linear combination of the FbSQL ones — so the
difference is representation, not estimation; standard errors, being
parameterization-dependent, differ accordingly. Around the model, the
observed behaviors were: the default pipeline raised an error when scoring
rows containing NULL; setting `handleInvalid = "skip"` silently dropped
those rows, returning three predictions for five input customers; and the
fitted model is a host-language `PipelineModel` object, with
`transform()` providing relation-in / relation-out shape at the DataFrame
level rather than in SQL.

## Overhead

A language layer should not cost more than it returns, so the evaluation
closes by bounding the price of the interface — a bound, not a
performance claim; statistical computing performance remains outside
FbSQL's scope. Table 3 reports median wall-clock times on the reference
environment for a deterministic synthetic churn dataset (companion
repository, script 13). Two observations follow. Fitting through
`fit_glm()` adds a roughly constant two hundred milliseconds to a plain R
`glm()` call at small sizes — the cost of invoking the embedded
interpreter and transferring the relation — and tracks the estimator
thereafter, staying under half a second end to end at $10^5$ rows.
Prediction, which runs entirely in SQL with no R, scores $10^5$ rows in
under a tenth of a second and scales linearly. At the analyst-scale
relations the proof of concept targets, the interface costs a fraction of
a second; large-scale estimation remains a matter for replacement engines
(see the *Discussion* section).

```{r tab3, echo=FALSE, results='asis'}
tab <- readLines(if (knitr::is_latex_output())
    "tables/overhead_benchmark.tex" else "tables/overhead_benchmark.md")
cat(tab[!grepl("^%", tab)], sep = "\n")
```

## Summary

The evaluation establishes three observations. First, *R compatibility*:
the reference implementation reproduces R's `glm()` and `predict.glm()`
exactly at the tested precision, in its own conformance suite and in an
independent re-run — thirteen of thirteen parity comparisons. Second,
*SQL language design*: the same statistical task, on systems that in two
cases fit numerically the same model, required materially different user
programs — manual encoding and coefficient arrays on MADlib, an
algorithm-and-deployment vocabulary on PostgresML, a host-language
pipeline with non-R formula semantics on Spark — confirming that the
differences at issue are design choices rather than capability gaps.
Third, *relation representation*: FbSQL was the only evaluated setting in
which the fitted model itself was a queryable relation, and the behaviors
observed around the edges of the workflow — silent reference-level
scoring, hard NULL errors, silently dropped rows, reparameterized
coefficient tables — each trace back to where a system keeps its model
and how the user must describe it. What these observations imply for the
trade-offs FbSQL accepted — the denormalized model relation, the JSONB
metadata, the column definition list — is the subject of the next
section.

# Discussion

The preceding sections presented FbSQL's specification and what its
evaluation made observable. This section steps back and examines the
design trade-offs behind those choices — what was given up, what was
gained, and where the boundaries of the present work lie. Nothing here is
a new result; it is the accounting that the *Design Principles* section
promised.

## One relation versus normalized relations

The seventeen-column model relation mixes two granularities: term-level
columns that vary by row, and model-level columns — `aic`, the
complete-case counters, `family`, and ultimately `metadata` — that repeat
the same value on every row. By the normalization discipline of the
relational tradition [@codd1972normalization], this is the textbook case
for decomposition: the model-level columns depend on the model, not the term,
and a normalized design would return a coefficient relation, a model
summary relation, and perhaps a metadata relation. We chose not to, and
the reason is closure. A SQL function returns one value; the moment
fitting returns two relations, the result can no longer flow through
`CREATE TABLE AS`, subqueries, and function composition as a single
thing, and the property the whole design exists to preserve is lost at
its most important boundary. Between normalization and closure, FbSQL
consistently chooses closure, and it pays deliberately: redundant values
that a hand-edited model table could in principle make inconsistent, and
a relation that carries more width than a term table strictly needs. The
practical costs are modest — the row count equals the number of terms, so
the repetition is small, and model-level values are one `SELECT DISTINCT`
away — but we present this as a trade-off, not a free lunch. The same
accounting applies to the `metadata` column: a separate metadata relation
would be cleaner by decomposition and was rejected on identical grounds.

## Metadata as part of the language

It would be easy to read the JSONB `metadata` column as an implementation
convenience — a place where `predict_glm()` stores what it happens to
need. We argue the opposite: metadata belongs to the language
specification, because without it the model relation does not mean
anything definite. A coefficient table alone under-determines the model.
`xlevels` records the level sets — including the reference level, which
appears in no `term` — without which `genderM` is an uninterpretable
label; `term_labels` preserves the formula's structure without relying on
re-parsing a string; `coef_terms` gives the canonical set of coefficients,
which is what makes an *unordered* collection of rows checkable for
completeness and duplication; `data_classes` fixes how each column was
interpreted. These fields are not there because the current predictor
consumes them — any consumer faces the same questions, whether it is a
predictor, an auditor asking what data the model saw, or a query
comparing two refits. They are there so that the relation is
self-describing: its meaning travels with it rather than residing in
session state or out-of-band knowledge. That is also why the schema is
versioned — an interpretation contract is an API, and `meta_version` is
its compatibility handle. The trade-off is acknowledged: a JSONB column is
a departure from flat relational form, adopted because the alternatives
sacrificed either closure (a second relation) or the relation's
granularity (flattening levels into extra rows), and because JSONB keeps
the contract queryable from SQL itself.

## SQL standard versus practical SQL

Relational theory offers a radical answer to SQL's accumulated
compromises: *The Third Manifesto* [@darwen1995third] judges SQL beyond
repair and specifies a new relational language, **D**, in which the
principles this paper cares about hold by construction. FbSQL's wager is
complementary rather than opposed — that the same principles can be
honored *inside* SQL as actually deployed, at the price of the explicit
trade-offs documented in this paper. That wager shapes our position on
syntax. FbSQL's normative call style,
`fit_glm(relation => ..., formula => ...)`, uses named notation that
PostgreSQL and Oracle provide but the ISO SQL standard does not. We are
deliberately cautious here. On one hand, the
experience of this paper suggests that named arguments are not cosmetic
for statistical interfaces: fitting functions accumulate optional
parameters, and the order-independence principle extends naturally from
rows to arguments. On the other hand, FbSQL is otherwise conservative —
everything else is plain functions over relations, installable as an
extension without any change to the SQL grammar — and we would not want
the proposal read as a criticism of the standard. The contrast with
BigQuery ML [@bigqueryml], which extends the SQL grammar itself with
model-creating DDL statements, is instructive as a design alternative:
grammar extension buys dedicated syntax at the cost of portability, while
the function route stays within what every SQL implementation already
parses. At most, we offer FbSQL's experience as one data point for future
standardization discussions: if statistical functions become common in
SQL, standardized named-argument notation would carry much of the
readability weight that dedicated syntax otherwise would.

## Beyond GLM

The *Minimum Atomic Relation* question — the smallest relation supporting
both interpretation and prediction — was answered for GLMs with the
term-grain coefficient relation. The question, not the answer, is the
reusable part, and it is worth asking conceptually for the tree ensembles
named throughout this paper. For a random forest, or a gradient-boosted
ensemble in the LightGBM or XGBoost families, the natural grain is the
*node*: one row per node per tree, with columns identifying the tree and
node, the split variable and threshold (or the level set, for categorical
splits), the child references or leaf value, and the treatment of missing
values at that split. Interpretation is then aggregation — variable
importance and split-frequency queries are `GROUP BY split_variable` away
— and prediction is traversal, reconstructible from the relation alone,
just as the linear predictor was reconstructible from coefficient rows.
Ensemble-level facts (number of trees, learning rate, sampling fractions)
repeat on every row exactly as `aic` does today, and the metadata
contract — level sets, column types, a canonical node inventory — carries
over with its role unchanged. We stress that this is a design argument,
not an implementation plan: the point is that nothing in the five
principles or in the relation-representation discipline is specific to
models that happen to be linear.

## Limitations

The present work has limits of two kinds, and we distinguish them
honestly. The first kind is *scope deliberately excluded from this paper*.
Offsets and observation weights are statistically routine and cheap to
add, and were left out to keep the proof of concept — and the reviewer's
task — minimal; the same holds for families beyond gaussian and binomial
and for non-canonical links. These are specification work already sketched
by the design, not open problems. The second kind is *genuine gaps in the
current specification and implementation*. Interactions are accepted at
fit time but rejected with an explicit error at prediction, because the
dummy-reconstruction contract does not yet cover product terms. Contrasts
are fixed to treatment coding; custom contrasts are unsupported.
Uncertainty reporting stops at Wald intervals for coefficients — there are
no prediction intervals, and for binomial models FbSQL returns
probabilities only, with no class-label prediction or `type` argument.
The `SETOF record` calling convention obliges the caller to write a column
definition list; removing that burden (through generated views, or an
implementation able to infer the output row type) is design work we have
not done. And the reference implementation is single-node: fitting passes
the data through an in-memory R engine, so FbSQL as implemented today says
nothing about large-scale estimation — by declared non-goal, but the
limitation is real for users with large relations.

## Future work

The directions that follow are research directions, not a roadmap. The
language–engine separation invites *alternative engines* — C, GPU, or
distributed implementations, including delegation to existing in-database
estimators such as MADlib's [@hellerstein2012madlib] — whose conformance
is in every case testable against the suite rather than arguable. The
estimator surface can grow to the remaining GLM families and then to the
node-grain tree-ensemble relations sketched above; the open language
questions are the call syntax a future standard could bless and an
output-typing mechanism that removes the column definition list. Finally,
a PGXN release and archived replication material place the language and
its evidence where their communities expect to find them.

# Conclusion

This paper proposed a language, not a package. FbSQL is a statistical
modeling DSL that takes SQL's own design principles — set orientation,
declarativeness, closure, order independence, and NULL semantics — as hard
constraints, and answers them with a formula as the model specification, a
relation as the fitted model, and prediction as a mapping from relations
to a relation. The PostgreSQL extension that accompanies the paper is the
reference implementation of that language, and the two `glm` functions are
its proof of concept; both exist to make the specification concrete and
testable, and neither is the claim. The estimator behind `fit_glm()` was
chosen precisely because it is ordinary.

Beyond the specification itself, the work yielded design lessons that we
believe outlast the proof of concept. First, SQL language design for
statistical modeling can be discussed — and verified — independently of
the execution engine: by fixing semantics by reference to an existing
system and locking them with a conformance suite, the specification became
a tested artifact, and prediction could be re-implemented in a different
language from the public relation alone. Second, the representation of the
fitted model is not an implementation detail but the heart of the language
question: the behavioral differences observed across systems in the
evaluation — what a user must build by hand, what fails, what proceeds
silently — traced back, case by case, to where each system keeps its model
and how the user must describe it. Third, a relation representation serves
prediction, interpretation, and audit with a single artifact: the same
queryable table that a predictor consumes also answers, by ordinary
`SELECT`, what the model is and what data it saw. And fourth, the
principles themselves can conflict — closure against normalization, most
prominently — so a design in this space is a set of explicit trade-offs;
what a language proposal owes its users is not the absence of such
choices but their documentation.

The question we most hope travels beyond this paper is the *Minimum
Atomic Relation* question: for a given class of statistical models, what
is the smallest relation that supports both the interpretation of a fitted
model and the reconstruction of its predictions? For generalized linear
models the answer proved to be a term-grain coefficient relation with
self-describing metadata; for random forests and gradient-boosted
ensembles it is plausibly a node-grain relation of split rules; for future
model classes it is a question each must answer before it can enter SQL
*as SQL*. Posed this way, statistical modeling in databases stops being
only a systems problem — how to compute efficiently near the data — and
becomes also a language problem: which properties of SQL must survive the
encounter, and what relational form statistical objects must take for that
to happen. The five principles and the running example of this paper are
offered as one worked answer and a template for the next ones.

The immediate continuations — alternative engines behind the unchanged
specification, further families and the tree-ensemble relations, and
public release of the extension and its replication material — were
discussed above and are underway; they extend the evidence, while the
proposal this paper stands on is the language.

# Replication material

Everything reported in this paper can be re-executed from two public
repositories whose division of labor is fixed. The **FbSQL repository**
(<https://github.com/dsc-chiba-u/FbSQL>) holds the extension source, the
conformance suite — which contains the running example verbatim, so every
number printed in the *Running Example* section is re-verified by
`make installcheck` — the Docker development environment, and this
manuscript. The **FbSQL-experiments repository**
(<https://github.com/dsc-chiba-u/FbSQL-experiments>) holds the comparative
material: pinned Docker environments for Apache MADlib, PostgresML, and
Spark MLlib, numbered scripts that reproduce the running example on each
system, the R-parity comparison, the per-claim sources for the two
literature-based systems, and the version-controlled comparison data
(`data/related_work.csv`) from which Table 1 and Table 2 of this paper are
generated by script — the paper never edits its tables by hand. All raw
outputs and summary tables are committed, so a re-run surfaces any
deviation as a diff. Scripts resolve repository locations through
environment variables or the sibling-checkout convention, with no
developer-specific paths.

<!-- TODO(replication): archive snapshots of both repositories at Zenodo
     at submission time and insert the DOIs here (fbrglm pattern). -->

# Computational details {.unnumbered}

The reference implementation was developed and tested with PostgreSQL
16.14 [@postgresql], PL/R 8.4.8.6 [@plr], and R 4.2.2 [@rcore], pinned in
the Docker image defined by `docker/Dockerfile` in the FbSQL repository;
no other runtime dependencies exist, and prediction requires no R. The
`pg_regress` conformance suite (eleven test files) is executed inside that
image by GitHub Actions on every commit. The comparison systems ran in
their own pinned Docker environments in the companion repository: Apache
MADlib 1.21.0 on PostgreSQL 11, PostgresML 2.7.12 on its bundled
PostgreSQL 15, and Spark MLlib 3.5.1 in local mode. This manuscript is
rendered from a single R Markdown source in a further pinned image
(`rocker/verse` 4.4.2), with its figures generated by committed R scripts
and its tables generated from the companion repository's data.

<!-- TODO(release sync): once the PGXN release of 0.1.0 is live, update
     the sentence below from "planned" to the released state (pgxn install
     fbsql) and re-check the README wording it references. -->

# Software availability {.unnumbered}

FbSQL is open-source software under the MIT license. The extension, its
conformance suite, and the Docker development environment are available
from the FbSQL repository on GitHub, whose address is given with the
replication material above; at the time of writing the extension installs
from a source checkout via PGXS (`make install`), as described in the
repository README. A release through PGXN, the PostgreSQL
Extension Network, is planned; the release metadata (`META.json`) is
already part of the repository.

# Acknowledgments {.unnumbered}

We thank the developers of PostgreSQL and of PL/R for the infrastructure
on which the reference implementation rests, and the communities behind
the open-source systems compared in this paper — Apache MADlib,
PostgresML, Apache Spark, Apache Hivemall, and H2O — whose released
software and documentation made the comparative evaluation possible. We
thank the Data Science Core, Chiba University, for institutional support,
and the RIKEN TRIP Headquarters Advanced General Intelligence for Science
(AGIS) Program for hosting the Omics AI Research Team during the
development of this work.

<!-- TODO(acknowledgments): confirm before submission — funding sources /
     grant numbers (none listed yet), coauthors' institutional
     acknowledgments (Juntendo / Chuo), and any individual reviewers or
     colleagues to be named. -->

# Declarations {.unnumbered}

**Funding** [To be completed at submission.]

**Competing interests** [To be completed at submission.]

**Data availability** All data used in this paper are small synthetic
datasets included in the public repositories: the running-example table
ships inside the extension's conformance suite and as `data/customer.csv`
in the companion repository, and the comparison data is version-controlled
there as `data/related_work.csv`.

**Code availability** The FbSQL extension and the companion experiment
repository are open source under the MIT license; see the replication
material section. [Zenodo DOIs to be added at submission.]

**Authors' contributions** [To be completed at submission.]

<!-- TODO(declarations): complete funding, competing interests, and
     author-contribution statements with all authors before submission
     (VLDBJ requires these). -->
